High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-10-17 doc/asplos2013/response.txt

We would like to thank the reviewers for the detailed feedback.  Below we address
the comments and questions most frequently raised by the reviewers.

Reviewers 1, 2, and 4 requested clarification of which components of the the API are
novel and which are similar to prior work.  Although events, locks, barriers, and reductions
have appeared in many forms, we believe the following contributions are novel:
  a) the use of generational events to efficiently describe the very large number of dependencies
that exist in a distributed memory application
  b) the deferred lock construct in which access to a critical resource can be mediated without
blocking operations that can waste execution resources
  c) the ability to modify a barrier's expected arrival count which allows nested task subdivision
without requiring a parent task to explicitly wait for subtasks to finish
  d) a general treatment of reductions that allows arbitrary reduction operations and improved
efficiency through specialized reduction instances that allow individual reduction operations to be
deferred and applied in bulk
  e) the ability to asychronously compose all of these constructs

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was use of the term "commutativity".  One traditional use of this term is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
can be composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the application of these
reduction operations to be deferred and reordered.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most reductions also satisfy our criterion of "foldability"
(e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
improve performance in most cases.

Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
to distinguish our implementation from those that are limited to shared-memory systems.  A more
accurate description would be "Distributed Memory Architectures."  We would be happy to change the 
title to clarify our meaning.

@@

Reviewer 2 noted that APIs like OpenCL supports events for asynchronously launching tasks and moving data. 
As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
single address space (e.g. node), whereas our events which are valid in every address space on 
distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
events and consequently cannot compose with other operations the way the synchronization mechanisms in our runtime can.

Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
the existing implementations on larger machine sizes.  All three reference implementations, two of which
were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
what portion of that speedup results from the composable, asynchronous nature of our 
runtime system.  To factor out any performance benefits gained by the Legion programming model 
independent from our runtime, we compared against bulk-synchronous versions of the applications from [4] 
written in Legion instead of the reference implementations.

Reviewer 3 commented that the experimental sections of the paper failed to evaluate the composability claim
of the runtime API.  It is unclear to us what kind of experiments would be useful to quantify the composability
of the API, but we would be open to suggests from the reviewers.  Our argument for compsability is a qualitative
one, centered around the ability of events to express dependences between all operations supported by the
runtime: computation, data movement, and synchronization.  As far as we are aware, no other API currently used
for parallel programming shares this property.

Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison
for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
Instead, we view the runtime presented in this paper as orthogonal to work on dataflow programming models.   
The runtime presented here provides a lower-level layer of abstraction that could serve as the 
foundation for many different dataflow programming systems.

Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
literature.  I-structures support a producer-consumer, write-once semantics that has some of the
properties of events in our runtime.  In contrast, deferred locks provide
unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
properties like serializability without the requirement of ordering.  We are unaware of any feature of
dataflow language implementations that support a similar semantics to deferred locks.

Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
are much lighter-weight, but current programming models do not allow these threads to be suspended.)
This cost is magnified by a programmer's need to conservatively create new threads whenever an API
call _might_ block or risk performance loss from the serialization of independent subtasks that are
launched from the same parent task.

Reviewers 1 and 5 asked for clarification regarding barriers and the need for a dynamic change
in arrival count.  In our experience, we've found that
the number of expected arrivals at a barrier is often not known at barrier creation-time (intuitively this
is necessary to support nested parallelism in Legion).  To support this paradigm, we added the ability to 
dynamically alter the arrival count.  To implement this feature on the distributed 
machines we're targetting, we use the well-studied vector clock algorithm from the distributed systems
literature (see citation [16] for additional details on the semantics and implementation of vector clocks).

Reviewers 1 and 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and each task
launched on a GPU processor results in an asyncrhonous CUDA grid launch.  GPU memories (framebuffer and zero-copy) 
are exposed in the same way as CPU main memory and the globally accessible "GASNet" memory.  Copies of 
data between GPU and other memories are explicitly requested by the programmer (like all other copies), 
and are performed with asynchronous DMA requests to the CUDA driver.  The CUDA driver provides device events, 
which are exposed in our runtime as general events that allow GPU operations to be composeable as well.

### 2012-10-17 doc/asplos2013/response.txt

Reviewer 3 commented that the experimental sections of the paper failed to evaluate the composability claim
of the runtime API.  It is unclear to us what kind of experiments would be useful to quantify the composability
of the API, but we would be open to suggests from the reviewers.  Our argument for compsability is a qualitative
one, centered around the ability of events to express dependences between all operations supported by the
runtime: computation, data movement, and synchronization.  As far as we are aware, no other API currently used
for parallel programming shares this property.

Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison
for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
Instead, we view the runtime presented in this paper as orthogonal to work on dataflow programming models.   
The runtime presented here provides a lower-level layer of abstraction that could serve as the 
foundation for many different dataflow programming systems.

Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
literature.  I-structures support a producer-consumer, write-once semantics that has some of the
properties of events in our runtime.  In contrast, deferred locks provide
unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
properties like serializability without the requirement of ordering.  We are unaware of any feature of
dataflow language implementations that support a similar semantics to deferred locks.

Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
are much lighter-weight, but current programming models do not allow these threads to be suspended.)
Instead our system provides lighter-weight tasks which the programmer
can launch at a much finer granularity without incurring the overhead of a thread.  Due to the minimal
overhead of tasks, programmers can express a much finer granularity of parallelism which our runtime
can then efficiently map onto available hardware resources.  Existing programming systems like MPI, OpenMP,
and P-threads entail too much overhead to leverage this finer-grained task parallelism.

Reviewer 1 and reviewer 5 asked for clarification regarding barriers and the need for a dynamic change
in arrival count.  In our experience, we've found that
the number of expected arrivals at a barrier is often not known at barrier creation-time (intuitively this
is necessary to support nested parallelism in Legion).  To support this paradigm, we added the ability to 
dynamically alter the arrival count.  To implement this feature on the distributed 
machines we're targetting, we use the well-studied vector clock algorithm from the distributed systems
literature (see citation [16] for additional details on the semantics and implementation of vector clocks).

Reviewer 1 and reviwer 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and each task
launched on a GPU processor results in an asyncrhonous CUDA grid launch.  GPU memories (framebuffer and zero-copy) 
are exposed in the same way as CPU main memory and the globally accessible "GASNet" memory.  Copies of 
data between GPU and other memories are explicitly requested by the programmer (like all other copies), 
and are performed with asynchronous DMA requests to the CUDA driver.  The CUDA driver provides device events, 
which are exposed in our runtime as general events that allow GPU operations to be composeable as well.

### 2012-10-17 doc/asplos2013/response.txt

We would like to thank the reviewers for the detailed feedback.  Below we address
the comments and questions most frequently raised by the reviewers.

Reviewers 1, 2, and 4 requested clarification of which components of the the API are
novel and which are similar to prior work.  Although events, locks, barriers, and reductions
have appeared in many forms, we believe the following contributions are novel:
  a) the use of generational events to efficiently describe the very large number of dependencies
that exist in a distributed memory application
  b) the deferred lock construct in which access to a critical resource can be mediated without
blocking operations that can waste execution resources
  c) the ability to modify a barrier's expected arrival count which allows nested task subdivision
without requiring a parent task to explicitly wait for subtasks to finish
  d) a general treatment of reductions that allows arbitrary reduction operations and improved
efficiency through specialized reduction instances that allow individual reduction operations to be
deferred and applied in bulk
  e) the ability to asynchronously compose all of these constructs

Reviewer 3 commented that the experimental sections of the paper failed to evaluate the composability claim
of the runtime API.  It is unclear to us what kind of experiments would be useful to quantify the composability
of the API, but we would be open to suggests from the reviewers.  Our argument for composability is a qualitative
one, centered around the ability of events to express dependences between all operations supported by the
runtime: computation, data movement, and synchronization.  As far as we are aware, no other API currently used
for parallel programming shares this property.

Reviewer 2 noted that APIs like OpenCL supports events for asynchronously launching tasks and moving data. 
As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
single address space (e.g. node), whereas our events are valid in every address space on 
distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
events and consequently cannot compose with other operations the way the synchronization mechanisms in our runtime can.

Reviewers 1 and 5 asked for clarification regarding barriers and the need for a dynamic change
in arrival count.  In programming models that support nested parallelism,
the number of expected arrivals at a barrier is often not known at barrier creation-time.
To allow tasks to be dynamically subdivided, we added the ability to 
alter the arrival count after the barrier has been created.  We use the well-studied vector 
clock algorithm from the distributed systems literature to
ensure that every node has observed all barrier alterations before completing the barrier, 
eliminating race conditions between alterations to and arrivals at the barrier.
(See citation [16] for additional details on the semantics and implementation of vector clocks.)

@@

Reviewers 1 and 5 asked for clarification regarding barriers and the need for a dynamic change
in arrival count.  In programming models that support nested parallelism,
the number of expected arrivals at a barrier is often not known at barrier creation-time.
To allow tasks to be dynamically subdivided, we added the ability to 
alter the arrival count after the barrier has been created.  We use the well-studied vector 
clock algorithm from the distributed systems literature to
ensure that every node has observed all barrier alterations before completing the barrier, 
eliminating race conditions between alterations to and arrivals at the barrier.
(See citation [16] for additional details on the semantics and implementation of vector clocks.)

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was use of the term "commutativity".  One traditional use of this term is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
can be composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the application of these
reduction operations to be deferred and reordered.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most reductions also satisfy our criterion of "foldability"
(e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold instances to further
improve performance in most cases.

Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
are much lighter-weight, but current programming models do not allow these threads to be suspended.)
This cost is magnified by a programmer's need to conservatively create new threads whenever an API
call might block.  Failing to do this can result in performance loss from the serialization of 
independent subtasks that are launched from the same parent task.

@@

Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison
for dataflow models would be the Legion programming system which is beyond the scope of this paper.  
Instead, we view the runtime presented in this paper as orthogonal to work on dataflow programming models. 
The runtime presented here provides a lower-level layer of abstraction that could serve as the 
foundation for many different dataflow programming systems.

Reviewer 1 noted that use "Distributed Architectures" in our title might be confusing.  Our goal was
to distinguish our implementation from those that are limited to shared-memory systems.  A more
accurate description would be "Distributed Memory Architectures."  We would be happy to change the 
title to clarify our meaning.

Reviewers 1 and 5 both inquired about how the runtime works with GPUs.  Our runtime uses CUDA to 
control the GPUs in the cluster.  A GPU is exposed as another kind of processor in our system, and each task
launched on a GPU processor results in an asynchronous CUDA grid launch.  GPU memories (framebuffer and zero-copy) 
are exposed in the same way as CPU main memory and the globally accessible "GASNet" memory.  Copies of 
data between GPU and other memories are explicitly requested by the programmer (like all other copies), 
and are performed with asynchronous DMA requests to the CUDA driver.  The CUDA driver provides device events, 
which are exposed in our runtime as general events that allow GPU operations to be composable as well.

### 2012-10-17 doc/asplos2013/response.txt

Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
the existing implementations on larger machine sizes.  All three reference implementations, two of which
were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
what portion of that speedup results from the composable, asynchronous nature of our 
runtime system.  To factor out any performance benefits gained by the Legion programming model 
independent from our runtime, we compared against bulk-synchronous versions of the applications from [4] 
written in Legion instead of the reference implementations.

Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
literature.  I-structures support a producer-consumer, write-once semantics that has some of the
properties of events in our runtime.  In contrast, deferred locks provide
unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
properties like serializability without the requirement of ordering.  We are unaware of any feature of
dataflow language implementations that support a similar semantics to deferred locks.

@@@ Version 1
Reviewer 3 commented that the experimental sections of the paper failed to evaluate the composability claim
of the runtime API.  It is unclear to us what kind of experiments would be useful to quantify the composability
of the API, but we would be open to suggests from the reviewers.  Our argument for composability is a qualitative
one, centered around the ability of events to express dependences between all operations supported by the
runtime: computation, data movement, and synchronization.  As far as we are aware, no other API currently used
for parallel programming shares this property.

@@@ Version 2
Reviewer 3 had concerns related to the composability of the constructs and the performance benefits of that
composability.  They key to the composability of our constructs is the event object.  Every operation in
our API (task launch, barrier arrival, lock request/release, data copy) accepts an event as a 'wait_for'
parameter, allowing that action to be deferred until some the provided event has occurred.  Every operation
provides, either directly or indirectly, an event that indicates when that operation has been completed.  By
linking operations together with events, a task can create an arbitrary dependency graph of operations to be
executed by the runtime, all asynchronously to the execution of the original task.  As clarified above, our results
section demonstrates that this decoupling of the dependency graph from the original requesting thread provides
significant speedup for applications written in Legion, especially for larger cluster sizes where the number of 
tasks and the latency of communication grow.

### 2012-10-18 doc/asplos2013/response.txt

Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
relative to hand-coded applications and whether the applications fit well into the bulk-synchronous model.  
The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
the existing implementations on larger machine sizes.  All three reference implementations, two of which
were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
what portion of that speedup results from the composable, asynchronous nature of our 
runtime system.  To factor out any performance benefits gained by the Legion programming model 
independent from our runtime, we compared against bulk-synchronous versions of the applications from [4] 
written in Legion instead of the reference implementations.

Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
literature.  I-structures support a producer-consumer, write-once semantics that has some of the
properties of events in our runtime.  In contrast, deferred locks provide
unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
properties like serializability without the requirement of ordering.  We are unaware of any feature of
dataflow language implementations that support a similar semantics to deferred locks.

Reviewer 3 had concerns related to the composability of the constructs and the performance benefits of that
composability.  They key to the composability of our constructs is the event object.  Unlike other 
parallel programming APIs, every operation in
our API (task launch, barrier arrival, lock/unlock, data copy) accepts an event as a 'wait_for'
parameter, causing that action to be deferred until the specified event has occurred.  With the exception of
'unlock' which completes immediately, every operation provides an event that indicates when that operation has been completed.  By
linking operations together with events, an arbitrary dependency graph of operations can be created and executed by the runtime
with no further interaction with the requesting task.  Without the composability of operations provided by events this would 
not be possible.  As our experimental results
demonstrate, this decoupling of the dependency graph from the original requesting task provides
significant speedup for applications written in Legion, especially when scaling to larger cluster sizes 
with increased communication latency.

### 2012-10-18 doc/asplos2013/response.txt

We would like to thank the reviewers for the detailed feedback.  Below we address
the comments and questions most frequently raised by the reviewers.

Reviewers 1, 2, and 4 requested clarification of which components of the the API are
novel and which are similar to prior work.  Although events, locks, barriers, and reductions
have appeared in many forms, we believe the following contributions are novel:
  a) the use of generational events to efficiently describe the very large number of dependencies
that exist in a distributed memory application
  b) the deferred lock construct in which access to a critical resource can be mediated without
blocking operations that can waste execution resources
  c) the ability to modify a barrier's expected arrival count which allows nested task subdivision
without requiring a parent task to explicitly wait for subtasks to finish
  d) a general treatment of reductions that allows arbitrary reduction operations and improved
efficiency through specialized reduction instances that allow individual reduction operations to be
deferred and applied in bulk
  e) the ability to asynchronously compose all of these constructs

Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
relative to hand-coded applications and whether bulk-synchronous implementations were appropriate 
for those applications.
The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
the existing implementations on larger machine sizes.  All three reference implementations, two of which
were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
what portion of that speedup results from the composable, asynchronous nature of our 
runtime system.  To factor out any performance benefits gained by the Legion programming model 
independent from our runtime, we compared against bulk-synchronous versions of the applications from [4] 
written in Legion instead of the reference implementations.

@@

Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
literature.  I-structures support a producer-consumer, write-once semantics that has some of the
properties of events in our runtime.  In contrast, deferred locks provide
unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
properties like serializability without the requirement of ordering.  We are unaware of any feature of
dataflow language implementations that support a similar semantics to deferred locks.

Reviewer 3 had concerns related to the composability of the constructs and the performance benefits of that
composability.  They key to the composability of our constructs is the event object.  Unlike other 
parallel programming APIs, every operation in our API (task launch, barrier arrival, lock/unlock, data copy)
accepts an event as a 'wait_for' parameter, causing that action to be deferred until the specified event has
occurred.  With the exception of 'unlock' which completes immediately, every operation also provides an event
that indicates when that operation has been completed.  By linking operations together with events, an arbitrary
dependency graph of operations can be created and then executed by the runtime with no further interaction with
the requesting task.  Without the composability of operations provided by events this would not be possible.
As our experimental results demonstrate, this decoupling of the dependency graph from the original requesting
task provides significant speedup for applications written in Legion, especially when scaling to larger cluster
sizes with increased communication latency.

@@

Reviewer 2 noted that APIs like OpenCL supports events for asynchronously launching tasks and moving data. 
As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
single address space (e.g. node), whereas our events are valid in every address space on 
distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
events and consequently cannot compose with other operations the way the synchronization mechanisms in
our runtime can.

Reviewers 1 and 5 asked for clarification regarding barriers and the need for dynamic changes
to a barrier's arrival count.  In programming models that support nested parallelism,
the number of expected arrivals at a barrier is often not known at the time the barrier is created.
To allow tasks to be dynamically subdivided, we added the ability to alter the arrival count after
the barrier has been created.  We use the well-studied vector clock algorithm from the distributed
systems literature to ensure that every node has observed all barrier alterations before completing
the barrier, eliminating race conditions between alterations to and arrivals at the barrier.
(See citation [16] for additional details on the semantics and implementation of vector clocks.)

### 2012-10-18 doc/asplos2013/response.txt

We would like to thank the reviewers for the detailed feedback.  Below we address
the comments and questions most frequently raised by the reviewers.

Reviewers 1, 2, and 4 requested clarification of which components of the the API are
novel and which are similar to prior work.  Although events, locks, barriers, and reductions
have appeared in many forms, we believe the following contributions are novel:
  a) the use of generational events to efficiently describe the very large number of dependencies
that exist in a distributed memory application
  b) the deferred lock construct in which access to a critical resource can be mediated without
blocking operations that can waste execution resources
  c) the ability to modify a barrier's expected arrival count which allows nested task subdivision
without requiring a parent task to explicitly wait for subtasks to finish
  d) a general treatment of reductions that permits arbitrary reduction operations and improved
efficiency through specialized reduction instances that allow individual reduction operations to be
deferred and applied in bulk
  e) the ability to asynchronously compose all of these constructs

@@

Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
literature.  I-structures support a producer-consumer, write-once semantics that has some of the
properties of events in our runtime.  In contrast, deferred locks provide
unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
properties like serializability without the requirement of ordering.  We are unaware of any feature of
dataflow language implementations that support a similar semantics to deferred locks.

Reviewer 3 had concerns related to the composability of the constructs and the performance benefits of that
composability.  They key to the composability of our constructs is the event object.  Unlike other 
parallel programming APIs, every operation in our API (task launch, barrier arrival, lock/unlock, data copy)
accepts an event as a 'wait_for' parameter, causing that action to be deferred until the specified event has
occurred.  With the exception of 'unlock' which completes immediately, every operation also provides an event
that indicates when that operation has been completed.  By linking operations together with events, an arbitrary
dependency graph of operations can be created and then executed by the runtime with no further interaction with
the requesting task.  Without the composability of operations provided by events this would not be possible.
As our experimental results demonstrate, this decoupling of the dependency graph from the original requesting
task provides significant speedup for applications written in Legion, especially when scaling to larger cluster
sizes with increased communication latency.

Reviewer 2 noted that APIs like OpenCL support events for asynchronously launching tasks and moving data. 
As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
single address space (e.g. node), whereas our events are valid in every address space on 
distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
events and consequently cannot compose with other operations the way the synchronization mechanisms in
our runtime can.

@@

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was use of the term "commutativity".  One traditional use of this term is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
can be composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the application of these
reduction operations to be deferred and reordered.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most reductions also satisfy our criterion of
"foldability" (e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold
instances to further improve performance in most cases.

Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
are much lighter-weight, but current programming models do not allow these threads to be suspended.)
This cost is magnified by a programmer's need to conservatively create new threads whenever an API
call might block.  Failing to do this can result in performance loss from the serialization of 
independent subtasks that are launched from the same parent task.

Reviewer 4 requested a comparison to existing dataflow programming models.  A more apt comparison for
dataflow models would be the higher-level Legion programming system which is beyond the scope of this paper.  
Instead, we view the runtime presented in this paper as orthogonal to work on dataflow programming models. 
The runtime presented here provides a lower-level layer of abstraction that can serve as the basis for
implementing systems with deferred execution, potentially including dataflow systems.

### 2012-10-18 doc/asplos2013/response.txt

Reviewer 3 expressed doubts about the performance of the Legion versions of the applications 
relative to hand-coded applications and whether bulk-synchronous implementations were appropriate 
for those applications.
The fourth citation in our paper [4] (which will appear at Supercomputing 2012) compares these same 
applications written in Legion to hand-coded reference implementations, and shows the Legion versions outperforming
the existing implementations on larger machine sizes.  All three reference implementations, two of which
were written by third parties, were coded in a bulk-synchronous style.  In this paper, we aimed to quantify 
what portion of that speedup results from the composable, asynchronous nature of our 
runtime system.  To factor out any performance benefits gained by the Legion programming model 
independent from our runtime, we compared against bulk-synchronous versions of the applications from [4] 
written in Legion instead of the reference implementations.

Reviewer 4 was unsure about the difference between our deferred locks and I-structures from the dataflow
literature.  I-structures support a producer-consumer, write-once semantics that has some of the
properties of events in our runtime.  In contrast, deferred locks provide
unordered atomic semantics, permitting multiple writes to an object and allowing programmers to describe
properties like serializability without the requirement of ordering.  We are unaware of any feature from 
dataflow language implementations that support a similar semantics to deferred locks.

@@

Reviewer 3 had concerns related to the composability of the constructs and the performance benefits of that
composability.  They key to the composability of our constructs is the event object.  Unlike other 
parallel programming APIs, every operation in our API (task launch, barrier arrival, lock/unlock, data copy)
accepts an event as a 'wait_for' parameter, causing that action to be deferred until the specified event has
occurred.  With the exception of 'unlock' which completes immediately, every operation also provides an event
that indicates when that operation has been completed.  By linking operations together with events, an arbitrary
dependency graph of operations can be created and then executed by the runtime with no further interaction with
the requesting task.  Without the composability of operations provided by events this would not be possible.
As our experimental results demonstrate, this decoupling of the dependency graph from the original requesting
task provides significant speedup for applications written in Legion, especially when scaling to larger cluster
sizes with increased communication latency.

Reviewer 2 noted that APIs like OpenCL support events for asynchronously launching tasks and moving data. 
As we noted in our related work, OpenCL events (and similar constructs in other APIs) are only valid in a 
single address space (e.g. node), whereas our events are valid in every address space on 
distributed memory machines.  Furthermore, the synchronization primitives in APIs like OpenCL do not use 
events and consequently cannot compose with other operations the way the synchronization mechanisms in
our runtime can.

Reviewers 1 and 5 asked for clarification regarding barriers and the need for dynamic changes
to a barrier's arrival count.  In programming models that support nested parallelism,
the number of expected arrivals at a barrier is often not known at the time the barrier is created.
To support this paradigm for higher-level programming models that target our runtime, 
we added the ability to alter the arrival count after
a barrier has been created.  We use the well-studied vector clock algorithm from the distributed
systems literature to ensure that every node has observed all barrier alterations before completing
the barrier, eliminating race conditions between alterations to and arrivals at the barrier.
(See citation [16] for additional details on the semantics and implementation of vector clocks.)

@@

Reviewers 1 and 5 had several questions about our handling of reductions.  One source of confusion
was use of the term "commutativity".  One traditional use of this term is limited to binary operations
in which both the left and right operands are of the same type.  Our implementation allows reductions
that have different left and right operand types (e.g. scaling of a vector by a scalar) and treats
each reduction operation as a unary function applied (atomically) to the current value (e.g. V *= 3
becomes (*3)(V) ).  When multiple reductions are applied to the same value, these unary functions
can be composed, and it's the commutativity of that composition
(e.g. [(*3) o (*2)](X) = [(*2) o (*3)](X) for all X) that allows the application of these
reduction operations to be deferred and reordered.  Nearly all "reasonable" reduction operations
that one might propose satisfy this form of commutativity.  Most reductions also satisfy our criterion of
"foldability" (e.g. [(*3) o (*2)](X) = (*6)(X) for all X), allowing the use of our reduction fold
instances to further improve performance in most cases.

Reviewer 1 commented that blocking APIs do not necessarily stall processors if there are enough 
runnable threads to keep the processor busy.  While this is true, the overhead of having many threads
to support this kind of computation is high due to the cost of software threads on CPUs.  (On GPUs, threads
are much lighter-weight, but current programming models do not allow these threads to be suspended.)
This cost is magnified by a programmer's need to conservatively create new threads whenever an API
call might block.  Failing to do this can result in performance loss from the serialization of 
independent operations that are launched from the same parent task.
